---
title: "Adding a New Module"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Adding a New Module}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    eval = FALSE
)
```

This checklist walks through how to add a new plotting module to **vizModules** so it matches the package's organization, documentation, and testing standards.

## Quick Checklist

- [ ] Pick the plot function you are wrapping and name your module accordingly:
  - For **dittoViz** functions: use `dittoViz_<PlotName>` (e.g., `dittoViz_ScatterPlot` for `dittoViz::scatterPlot`)
  - For **plotthis** functions: use `plotthis_<PlotName>` (e.g., `plotthis_AreaPlot` for `plotthis::AreaPlot`)
  - For **custom/standalone** functions: use the plot name directly (e.g., `linePlot`, `piePlot`)
- [ ] If adding a brand-new plotting function (e.g., `piePlot`), define and document that plot function first, then wrap it with the module.
- [ ] Create the three core files: `R/<moduleName>_module_ui.R`, `R/<moduleName>_module_server.R`, and `R/<moduleName>_module_app.R`.
- [ ] Document UI, server, and app functions with roxygen (`@export`, params, examples).
- [ ] In the UI function docstring, add **three required sections**:
  - `@section Plot parameters not implemented or with altered functionality:` - List inputs not exposed and why
  - `@section Plot parameters and defaults:` - Document all exposed parameters with UI labels and defaults
  - `@section Plot parameters implementing new functionality:` - Document any new or plotly-specific controls (axes, ticks, reference lines, etc)
- [ ] Add an example app that uses the module twice to prove multi-instance behavior.
- [ ] Cover the base plotting function with `testthat`; cover the module/app with `shinytest2`.
- [ ] Note any `ggplotly` conversion quirks that change or drop functionality.
- [ ] **Never use `eval(str2expression())` on raw user input.** Use `safe_eval_filter()`, `validate_expression()`, or `safe_resolve_adj_fxn()` instead (see [Sanitizing User-Provided Expressions] below).

## Naming & Organization

- [ ] File names follow the module naming pattern:
  - **dittoViz wrappers**: `dittoViz_<PlotName>_module_ui.R` (e.g., `dittoViz_scatterPlot_module_ui.R`)
  - **plotthis wrappers**: `plotthis_<PlotName>_module_ui.R` (e.g., `plotthis_AreaPlot_module_ui.R`)
  - **Custom modules**: `<plotName>_module_ui.R` (e.g., `linePlot_module_ui.R`)
- [ ] Function names follow the pattern: `<moduleName>InputsUI()`, `<moduleName>OutputUI()`, `<moduleName>Server()`, `<moduleName>App()`.
  - Examples: `plotthis_AreaPlotInputsUI()`, `dittoViz_scatterPlotServer()`, `linePlotApp()`
- [ ] Internal helpers stay in existing helper files when broadly useful; otherwise keep them inside the module file.

## Documentation Standards

- [ ] Roxygen headers include title, description, parameters, return value, authors, and `@export`.
- [ ] Examples show minimal, runnable usage with a small dataset.
- [ ] **The UI function must include three documentation sections:**

### 1. `@section Plot parameters not implemented or with altered functionality:`

List all parameters from the base plot function that are **not** exposed via UI inputs, with explanations:

```r
#' @section Plot parameters not implemented or with altered functionality:
#' The following [plotthis::AreaPlot()] parameters are not available via UI inputs:
#' \itemize{
#'   \item \code{xlab} - X-axis label (plotly allows interactive editing)
#'   \item \code{ylab} - Y-axis label (plotly allows interactive editing)
#'   \item \code{title} - Plot title (plotly allows interactive editing)
#'   \item \code{subtitle} - Plot subtitle (not supported in plotly)
#'   \item \code{legend.position} - Legend positioning (plotly allows interactive repositioning)
#'   \item \code{split_by} - Split variable (returns a patchwork object, not supported in plotly)
#'   \item \code{palette} - Managed internally via the palette selection UI
#' }
```

### 2. `@section Plot parameters and defaults:`

Document all parameters that **are** exposed, listing their UI label and default value:

```r
#' @section Plot parameters and defaults:
#' The following [plotthis::AreaPlot()] parameters can be accessed via UI inputs and/or the \code{defaults} argument:
#' \itemize{
#'   \item \code{x} - X-axis variable (UI: "X values", default: 2nd categorical variable)
#'   \item \code{y} - Y-axis variable (UI: "Y values", default: 2nd numeric variable)
#'   \item \code{group_by} - Grouping variable (UI: "Group by", default: 3rd categorical variable or "")
#'   \item \code{facet_by} - Faceting variable (UI: "Facet by", default: "")
#'   \item \code{theme} - ggplot2 theme (UI: "Theme", default: "theme_this")
#'   \item \code{alpha} - Area fill transparency (UI: "Alpha", default: 1)
#' }
```

### 3. `@section Plot parameters implementing new functionality:`

Document all module-specific parameters (plotly controls, reference lines, etc.):

```r
#' The following parameters implementing new functionality or controlling plotly-specific features are also available:
#' \itemize{
#'   \item \code{axis.font.size} - Axis title font size (UI: "Axis font size", default: 18)
#'   \item \code{axis.showline} - Show axis border lines (UI: "Show axis lines", default: TRUE)
#'   \item \code{axis.tickfont.size} - Size of tick labels (UI: "Tick label size", default: 12)
#'   \item \code{hline.intercepts} - Y-coordinates for horizontal reference lines (UI: "Y-intercepts", default: "")
#'   \item \code{hline.colors} - Colors for horizontal lines, comma-separated (UI: "Colors", default: "#000000")
#'   \item \code{hline.linetypes} - Line types for horizontal lines, comma-separated (UI: "Line types", default: "dashed")
#'   \item \code{vline.intercepts} - X-coordinates for vertical reference lines (UI: "X-intercepts", default: "")
#'   \item \code{abline.slopes} - Slopes for diagonal reference lines (UI: "Slopes", default: "")
#' }
```

**Note:** Reference line parameters (`hline.*`, `vline.*`, `abline.*`) accept comma-separated values to control each line individually.

## Functionality & Non-Exposed Inputs

- [ ] For each plot function argument, decide: expose, set a fixed default, or drop.
- [ ] If dropped or fixed, document it inside the UI function (description + reason).
- [ ] If `ggplotly` alters or drops a feature (e.g., certain geoms, annotations), note that limitation in the UI docs so users know what to expect.

## Example App Requirement

- [ ] Provide an app in `<plot>_module_app.R` that instantiates the module **twice** with different datasets/IDs.
- [ ] Keep the app minimal: load sample data, render both modules' inputs/outputs, and no extra custom logic beyond demonstrating the module.
- [ ] Add the module to the gallery app (`inst/apps/module-gallery/app.R`), placing it in its own tab alongside the other modules.

## Testing Requirements

- [ ] `testthat`: cover the base plotting function's core behavior and arguments (data handling, grouping, palette handling, etc.).
- [ ] `shinytest2`: cover the module/app (rendering inputs, updating outputs, download buttons if present).
- [ ] Place tests under `tests/testthat/` with clear file names (`test-<plot>.R`, `test-<plot>-app.R`).
- [ ] Ensure tests run headless and deterministically (seed randomness where needed).
- [ ] For new plotting functions, add dedicated `testthat` coverage of the plotting helper itself (input validation, defaults, edge cases) in addition to the module tests.

## Implementing a New Plotting Function (e.g., `piePlot`)

- [ ] Add the plotting function under `R/` with full roxygen docs, inputs/returns, and examples.
- [ ] Keep arguments consistent with existing plot functions (data first, `...` last, palette/palcolor patterns).
- [ ] Document any assumptions about input shape (e.g., pre-summarized table for pies).
- [ ] Add `testthat` coverage for the plotting function (happy paths + invalid inputs).
- [ ] Only after the plotting function is stable, build the module UI/server/app wrappers around it.

## Gallery App

- [ ] Add or update the gallery app at `inst/apps/module-gallery/app.R` to include the new module in its own tab.
- [ ] Each tab should load a small sample dataset and show the module's inputs and outputs together.
- [ ] Verify namespacing: each module instance should have a unique `id` and independent state.
- [ ] Keep dependencies minimal (prefer built-in datasets) so the gallery runs out-of-the-box.

## Review Before Submitting

- [ ] Run `devtools::document()` to update NAMESPACE and Rd files.
- [ ] Run `devtools::check()` and ensure tests pass locally.
- [ ] Confirm UI text/tooltips mention any missing or altered plot features.
- [ ] Verify both module instances in the example app work independently (namespacing correct).

## Sanitizing User-Provided Expressions

**Never use `eval(str2expression())` or `eval(parse())` on raw user input.**
If a Shiny app is deployed publicly, this allows arbitrary code execution on the
server (e.g., `system("rm -rf /")`). VizModules provides three exported helper
functions for safely handling user-typed expressions. Use them whenever your module
accepts free-text input that will be evaluated or passed to a plotting function.

### `safe_eval_filter(expr_text, data)`

Use when a module **evaluates** a user-typed filter expression directly to produce
a logical vector for row subsetting. The expression is parsed, its AST is walked
to ensure only allowed operations are present (comparisons, logical operators,
column references, and literals), and then it is evaluated in a restricted
environment containing only the data frame's columns.

```{r}
# In a module server — filtering rows by a textInput:
rows.use = safe_eval_filter(isolate_fn(input$rows.use), data())
```

Returns a logical vector (same length as `nrow(data)`), or `NULL` if the input is
empty, unparseable, or contains disallowed operations.

### `validate_expression(expr_text, col_names)`
Use when a module **passes** a user-typed expression string through to a
downstream plotting function that will evaluate it internally (e.g.,
`plotthis::BoxPlot(highlight = ...)`). The string is validated but not executed.

```{r}
# In a module server — passing a highlight expression to plotthis:
highlight <- validate_expression(isolate_fn(input$highlight), names(data()))
```

Returns the original string if safe, or `NULL`.

### `safe_resolve_adj_fxn(fn_name)`

Use when a module resolves a function name from a dropdown or text input into an
actual function reference (e.g., for `x.adj.fxn`, `y.adj.fxn`). Only function
names in the allowed list (`"log2"`, `"log"`, `"log10"`, `"neg_log10"`, `"log1p"`,
`"as.factor"`, `"abs"`, `"sqrt"`) are accepted.

```{r}
# In a module server — resolving an adjustment function:
x.adj.fxn = safe_resolve_adj_fxn(isolate_fn(input$x.adj.fxn))
```

Returns the function, or `NULL` if the name is empty or not in the allowed list.

### What counts as "allowed"?

All three helpers share the same whitelist of safe AST nodes:

- **Comparisons:** `<`, `>`, `<=`, `>=`, `==`, `!=`
- **Logical operators:** `&`, `&&`, `|`, `||`, `!`
- **Utilities:** `%in%`, `c()`, `is.na()`, `is.null()`
- **Arithmetic:** `-`, `+`, `*`, `/`, `:`, `%%`
- **Grouping:** `()`
- **Column names** from the data
- **Literals:** numbers, strings, `TRUE`, `FALSE`, `NA`, `NULL`, `Inf`, `NaN`

Anything outside this list (including function calls like `system()`, `file.remove()`,
`library()`, etc.) is rejected and a warning is issued.
