---
title: "Understanding Colour Picker Inputs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Understanding Colour Picker Inputs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    eval = FALSE
)
```

## Introduction

VizModules uses colour picker inputs extensively to allow users to customize plot aesthetics.
This vignette explains the two types of colour pickers used in the package, their output formats,
and how to work with them in your Shiny modules.

## Types of Colour Pickers

VizModules uses two different colour picker implementations:

1. **`colourpicker::colourInput()`** - For single colour selection
2. **`multiColorPicker()`** - For selecting colours for multiple discrete groups

## Single Colour Picker: `colourpicker::colourInput()`

### Usage

The `colourpicker::colourInput()` widget is used throughout VizModules for selecting individual colours
such as axis line colors, title colors, annotation colors, and gradient endpoints.

```{r colourpicker-usage}
library(shiny)
library(colourpicker)

ui <- fluidPage(
    colourInput("my_color", "Choose a color", value = "#F0E442"),
    textOutput("color_value")
)

server <- function(input, output, session) {
    output$color_value <- renderText({
        paste("Selected color:", input$my_color)
    })
}
```

### Output Format

The colour picker returns a **character string** in **hexadecimal (HEX) format**:

- **Format**: `"#RRGGBB"` where RR, GG, BB are two-digit hexadecimal values (00-FF)
- **Examples**: `"#F0E442"`, `"#0072B2"`, `"#000000"`, `"#FFFFFF"`
- **Case**: Can be uppercase or lowercase (VizModules normalizes to uppercase)

### Accessing in Server Code

In server functions, access the colour value via `input$<inputId>`:

```{r accessing-color}
# Basic access
my_color <- input$my_color  # Returns "#F0E442"

# With isolation (recommended in VizModules)
isolate_fn <- setup_auto_update_logic(input)
my_color <- isolate_fn(input$my_color)

# Direct use in plotting
plot(x, y, col = input$my_color)
```

### Common Use Cases in VizModules

VizModules uses `colourInput()` for:

1. **Gradient colors** (scatter plots with continuous color scales):
   ```{r gradient-example}
   colourpicker::colourInput(ns("min.color"), "Minimum color", value = "#F0E442")
   colourpicker::colourInput(ns("max.color"), "Maximum color", value = "#0072B2")
   
   # In server:
   min.color = isolate_fn(input$min.color)  # "#F0E442"
   max.color = isolate_fn(input$max.color)  # "#0072B2"
   ```

2. **Axis and title styling**:
   ```{r axis-styling}
   colourInput(ns("text.colour"), "Title Color", value = "#000000")
   colourInput(ns("axis.linecolor"), "Axis Line Color", value = "black")
   colourInput(ns("axis.tickcolor"), "Tick Mark Color", value = "black")
   
   # In server:
   text.colour = isolate_fn(input$text.colour)          # "#000000"
   axis.linecolor = isolate_fn(input$axis.linecolor)    # "black"
   ```

3. **Plot-specific colors**:
   ```{r plot-colors}
   # Contour color for density overlays
   colourInput(ns("contour.color"), "Contour color", value = "black")
   
   # Line of best fit
   colourInput(ns("line.best.colour"), "Line of best fit colour", value = "#000000")
   
   # Pie chart elements
   colourInput(ns("slice.border.color"), "Slice Border Color", value = "#FFFFFF")
   ```

### Updating Programmatically

You can update colour pickers from the server using `colourpicker::updateColourInput()`:

```{r update-color}
# Reset to default value
colourpicker::updateColourInput(session, "min.color", value = "#F0E442")

# Change based on user action
observeEvent(input$preset_palette, {
    if (input$preset_palette == "warm") {
        colourpicker::updateColourInput(session, "my_color", value = "#FF6B6B")
    } else if (input$preset_palette == "cool") {
        colourpicker::updateColourInput(session, "my_color", value = "#4ECDC4")
    }
})
```

## Multi-Group Colour Picker: `multiColorPicker()`

### Usage

The `multiColorPicker()` widget is a custom VizModules component for selecting colours for multiple
discrete groups simultaneously. It's primarily used for categorical color mappings in scatter plots
and other plots that color by discrete variables.

```{r multicolor-usage}
library(VizModules)

# In UI
multiColorPicker(
    ns("color.panel"),
    label = "Color palette",
    groups = c("setosa", "versicolor", "virginica"),
    palette_options = default_palettes()[["choices"]],
    selected_palette = "dittoColors",
    colors = NULL,  # Optional: pre-set colors as named vector
    compact = TRUE
)
```

### Output Format

The multi-group colour picker returns a **named character vector** where:

- **Names**: The group/category names
- **Values**: HEX color codes for each group
- **Format**: `c(group1 = "#RRGGBB", group2 = "#RRGGBB", ...)`

**Example output:**
```{r multicolor-output}
# Example: Species coloring for iris data
color_mapping <- input$color.panel
# Returns:
# c(setosa = "#E69F00", versicolor = "#56B4E9", virginica = "#009E73")

# Accessing individual colors
setosa_color <- color_mapping["setosa"]  # "#E69F00"
```

### Accessing in Server Code

```{r accessing-multicolor}
# In dittoViz_ScatterPlot module:
color.panel <- reactive({
    isolate_fn <- setup_auto_update_logic(input)
    levels <- color_levels()  # Get current color groups
    
    # Access the multiColorPicker values
    picker_values <- isolate_fn(input$color.panel)
    
    # Returns a named vector like:
    # c(GroupA = "#E69F00", GroupB = "#56B4E9", GroupC = "#009E73")
    
    picker_values
})

# Use in plotting
dittoViz::scatterPlot(
    data(),
    color.by = "Species",
    color.panel = unname(color.panel()),  # Pass unnamed vector to dittoViz
    colors = seq_len(length(color.panel()))
)
```

### Dynamic Group Updates

The `multiColorPicker` automatically updates when the groups change (e.g., when a different
categorical variable is selected for coloring):

```{r dynamic-groups}
# Reactive to detect color groups
color_levels <- reactive({
    df <- data()
    color_by <- input$color.by
    
    if (is.null(df) || is.null(color_by) || color_by == "") {
        return(character(0))
    }
    
    # For factors, preserve order; for character, alphabetize
    if (is.factor(df[[color_by]])) {
        levels(df[[color_by]])
    } else {
        levels(as.factor(df[[color_by]]))
    }
})

# Render UI with current groups
output$color.panel.ui <- renderUI({
    groups <- color_levels()
    
    if (length(groups) == 0) {
        # No discrete groups - fall back to single color picker
        colourpicker::colourInput(ns("single.point.color"), 
                                   label = "Point color",
                                   value = "#000000")
    } else {
        # Show multi-group picker
        multiColorPicker(
            ns("color.panel"),
            label = "Color palette",
            groups = groups,  # Dynamic groups
            palette_options = default_palettes()[["choices"]],
            selected_palette = "dittoColors"
        )
    }
})
```

## Uniform Color Inputs

VizModules provides uniform color inputs across all modules via the `.uniform_axes_inputs_ui()`
and `.uniform_lines_inputs_ui()` helper functions defined in `uniform_ui_inputs.R`.

These helpers ensure consistent styling options across all plot types:

```{r uniform-inputs}
# From .uniform_axes_inputs_ui():
colourInput(ns("text.colour"), "Title Color", value = "#000000")
colourInput(ns("axis.title.font.color"), "Axis Title Color", value = "#000000")
colourInput(ns("axis.linecolor"), "Axis Line Color", value = "black")
colourInput(ns("axis.tickcolor"), "Tick Mark Color", value = "black")
colourInput(ns("axis.tickfont.color"), "Tick Label Color", value = "black")

# From .uniform_lines_inputs_ui():
colourInput(ns("line.best.colour"), "Line of best fit colour", value = "#000000")
```

All of these return standard HEX color strings when accessed via `input$<id>`.

## Practical Examples

### Example 1: Custom Color Scheme for Scatter Plot

```{r example1}
library(VizModules)
library(shiny)

ui <- fluidPage(
    sidebarLayout(
        sidebarPanel(
            dittoViz_ScatterPlotInputsUI("scatter", iris,
                defaults = list(
                    x.by = "Sepal.Length",
                    y.by = "Sepal.Width",
                    color.by = "Species"
                )
            )
        ),
        mainPanel(
            dittoViz_ScatterPlotOutputUI("scatter")
        )
    )
)

server <- function(input, output, session) {
    # Provide manual colors as a named vector
    manual_colors <- c(
        setosa = "#E31A1C",
        versicolor = "#1F78B4",
        virginica = "#33A02C"
    )
    
    dittoViz_ScatterPlotServer("scatter", 
                                data = reactive(iris),
                                manual.colors = manual_colors)
}

shinyApp(ui, server)
```

### Example 2: Programmatic Color Updates

```{r example2}
server <- function(input, output, session) {
    # Update colors based on a theme selector
    observeEvent(input$theme, {
        if (input$theme == "dark") {
            # Dark theme colors
            colourpicker::updateColourInput(session, "text.colour", value = "#FFFFFF")
            colourpicker::updateColourInput(session, "axis.linecolor", value = "#CCCCCC")
        } else {
            # Light theme colors
            colourpicker::updateColourInput(session, "text.colour", value = "#000000")
            colourpicker::updateColourInput(session, "axis.linecolor", value = "#000000")
        }
    })
}
```

### Example 3: Extracting Colors from multiColorPicker

```{r example3}
# In your module server:
observe({
    colors <- input$color.panel
    
    # Colors is a named character vector
    # c(GroupA = "#E69F00", GroupB = "#56B4E9", ...)
    
    # Get color for specific group
    group_a_color <- colors["GroupA"]
    
    # Get all colors as unnamed vector
    color_values <- unname(colors)
    
    # Get all group names
    group_names <- names(colors)
    
    print(paste("Groups:", paste(group_names, collapse = ", ")))
    print(paste("Colors:", paste(color_values, collapse = ", ")))
})
```

## Summary

- **`colourpicker::colourInput()`**: Returns a single HEX color string (`"#RRGGBB"`)
  - Used for: single colors, gradients, axis styling, annotations
  - Accessed via: `input$<id>` returns character string
  - Updated via: `colourpicker::updateColourInput(session, id, value)`

- **`multiColorPicker()`**: Returns a named character vector (`c(group1="#...", group2="#...")`)
  - Used for: discrete group color mappings in categorical plots
  - Accessed via: `input$<id>` returns named character vector
  - Updated automatically: when groups change, widget re-renders with new groups

Both types integrate seamlessly with R plotting functions that expect color parameters.
The HEX format is universally supported by base R graphics, ggplot2, plotly, and all
major visualization packages.

## See Also

- `vignette("custom-modules")` - Building wrapper modules with custom inputs
- `vignette("quick-start")` - Getting started with VizModules
- `?colourpicker::colourInput` - Documentation for the colourpicker package
- `?VizModules::multiColorPicker` - VizModules custom color picker
