---
title: "Building Custom Modules"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Building Custom Modules}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    eval = FALSE
)
```

## Introduction
 
The modules in **VizModules** are designed to be composed and extended.
You can build higher-level modules that add custom logic such as data filtering, transformations,
or additional UI controls while reusing the full functionality of the base modules.

This vignette demonstrates how to create a custom module by building on top of the `scatterPlot` module.

If you are working with an AI coding agent, the package ships an
[Agent Skill](https://agentskills.io) covering exactly this material -
**`vizmodules-custom-module`** - along with the namespace contract, reactive `defaults`,
avoiding double renders, manual-edit persistence, and model-line backends. Install it (and
its two siblings) into your project with `VizModules::use_vizmodules_skills(".")`, or pass
`client = "copilot"` / `client = "claude"` to write to `.github/skills/` or `.claude/skills/`
instead of the default `.agents/skills/`. See the README for details.

## The Pattern

When building a custom module, you need to handle Shiny's namespacing correctly. The key insight is:

1. **Your module's custom inputs** need to be namespaced with `NS(id)`.
2. **The base module's UI and server functions** should receive the bare `id`, not a namespaced version.
3. **Data processing** that requires access to your module's inputs must happen inside a `moduleServer()` block.
4. **The base module's server** should be called *outside* that `moduleServer()` block to avoid double-namespacing issues.

## A Minimal Example

Let's build a custom module that adds a simple filtering checkbox to the `scatterPlot` module.

### The UI

```{r module-ui}
library(VizModules)

minimalModuleUI <- function(id) {
    ns <- NS(id)
    tagList(
        h4("Minimal Module Controls"),
        # Custom input - uses the module's namespace
        checkboxInput(ns("filter_setosa"), "Start with Setosa Only", value = FALSE),
        hr(),
        # Base module UI - pass the bare 'id', not ns(id)
        dittoViz_scatterPlotInputsUI(id, iris)
    )
}

minimalModuleOutput <- function(id) {
    # Simply delegate to the base module's output UI
    dittoViz_scatterPlotOutputUI(id)
}
```

Notice that `checkboxInput()` uses `ns("filter_setosa")` to namespace the custom input,
while `dittoViz_scatterPlotInputsUI()` receives the bare `id`. This ensures the base module creates its inputs in the correct namespace.

### The Server

```{r module-server}
minimalModuleServer <- function(id, data_reactive) {
    # Step 1: Process data inside a moduleServer block
    # This gives us access to inputs namespaced to 'id' (our module's inputs)
    filtered_data <- moduleServer(id, function(input, output, session) {
        reactive({
            req(data_reactive())
            df <- data_reactive()

            # Input specific to this custom module
            if (isTRUE(input$filter_setosa)) {
                if ("Species" %in% names(df)) {
                    df <- df[df$Species == "setosa", ]
                }
            }
            df
        })
    })

    # Step 2: Call the base module server OUTSIDE the moduleServer block
    # This is critical! If we called this inside the moduleServer above,
    # dittoViz_scatterPlotServer would look for inputs at id-id-inputName instead of id-inputName
    dittoViz_scatterPlotServer(id, filtered_data)
}
```

**Why this pattern?**

- `moduleServer(id, ...)` gives us access to `input$filter_setosa`, which is namespaced to our wrapper's `id`.
- By calling `dittoViz_scatterPlotServer(id, filtered_data)` *outside* the `moduleServer()` closure, the base module attaches to the same namespace as our UI, not a nested one.
- If we called `dittoViz_scatterPlotServer()` inside the `moduleServer()` block, it would create nested namespaces like `id-id-x_axis`, which wouldn't match the actual input IDs in the UI.

### Putting It Together

```{r full-app}
ui <- fluidPage(
    titlePanel("Minimal Module Example"),
    sidebarLayout(
        sidebarPanel(
            minimalModuleUI("demo")
        ),
        mainPanel(
            minimalModuleOutput("demo")
        )
    )
)

server <- function(input, output, session) {
    # Pass a reactive data source to the module
    minimalModuleServer("demo", reactive({
        iris
    }))
}

shinyApp(ui, server)
```

## Hiding Base Module Inputs

If your module pre-sets certain parameters, you can hide those inputs from the user to keep them from being changed:

```{r hiding-inputs}
focusedModuleUI <- function(id) {
    ns <- NS(id)
    tagList(
        h4("Simplified Scatter Plot"),
        # The UI takes no hide.* arguments -- see the server below.
        dittoViz_scatterPlotInputsUI(id, iris)
    )
}

focusedModuleServer <- function(id, data_reactive) {
    # hide.inputs and hide.tabs are arguments of the *Server*, not the *InputsUI.
    dittoViz_scatterPlotServer(id, data_reactive,
        hide.inputs = c("shape.by", "color.by")
    )
}
```

## Driving Base Module Inputs From Your Wrapper

Hiding an input fixes it to one value. When a base module parameter instead needs
to *follow* something your wrapper computes, pass a `reactive()` as that entry of
`defaults` rather than reaching for `update*Input()`. The key must name an input the
module actually reads -- `get_default()` falls back silently, so a `defaults` entry for
an unexposed key does nothing at all. (`main` is the one to watch: no module exposes a
plot title, so `defaults = list(main = ...)` has no effect.)

```{r reactive-defaults}
colouredModuleServer <- function(id, data_reactive) {
    colour_col <- moduleServer(id, function(input, output, session) {
        reactive(if (isTRUE(input$filter_setosa)) "" else "Species")
    })

    # The colour mapping tracks the checkbox, but the user can still change it.
    dittoViz_scatterPlotServer(
        id, data_reactive,
        defaults = list(color.by = colour_col)
    )
}
```

This is the recommended parent-to-child channel for parameter values. It resolves
server-side, so the plot renders once per change instead of twice, and the control
stays user-editable. See `vignette("defaults-and-hiding", package = "VizModules")`
for the full semantics.

## Automatically Preserving Manual Title/Legend/Annotation Edits

The base modules let users interactively drag and edit the plot title, legend,
annotations, draggable axis titles, and continuous-colour legend (colorbar), and
they re-apply those manual tweaks across re-renders. When you wrap a base
module (as in the examples above), this persistence is inherited for free.

If you build a brand-new `plotly` output from scratch inside a custom
module (rather than delegating to a base module's server), you can add the same
behaviour simply with 3 basic steps:

```{r manual-edits}
customPlotServer <- function(id, data_reactive) {
    moduleServer(id, function(input, output, session) {
        # 1. A unique event source + the edit store, created once.
        plot_source <- session$ns("customplot")
        edit_store <- setup_manual_edits(input, session, plot_source)

        output$plot <- renderPlotly({
            # 2. Create your plotly object with your plotting function
            fig <- build_my_plotly_figure(data_reactive(), input)
            # 3. Restore + capture edits on every render, then return the figure.
            finalize_manual_edits(fig, plot_source, edit_store, session)
        })
    })
}
```

See the "Adding a New Module" vignette for the full description of these helpers, 
but this is pretty much all you need to do.

## Best Practices

When you add a free-text control of your own, debounce it. `textInput()` reports
on every keystroke, so a base module reading it directly is rebuilt once per
character — and if the text is an expression, most of those characters describe
something that cannot parse yet:

```{r debounce-wrapper}
myModuleServer <- function(id, data) {
    # Once, in the server body. 500-800ms is a good range.
    query <- debounce(reactive(input$query), 700)

    filtered <- moduleServer(id, function(input, output, session) {
        reactive({
            keep <- safe_eval_filter(query(), data())
            if (is.null(keep)) data() else data()[keep, , drop = FALSE]
        })
    })

    dittoViz_scatterPlotServer(id, filtered)
}
```

Select, numeric, and checkbox inputs need no such treatment — they report
discrete choices, not keystrokes. See the "Debouncing Free-Text Inputs" section
of `vignette("adding-a-new-module", package = "VizModules")` for the full
rationale.


1. **Keep wrapper logic focused**: Each wrapper should add a cohesive set of related functionality.

2. **Document the data requirements**: If your wrapper expects certain columns, a specific class or data structure, etc, document this clearly.

3. **Use reactive expressions**: Use reactive data inputs.
 
4. **Test the namespace**: If inputs aren't working, check that you're handling namespaces correctly.
A common symptom of namespace issues is that inputs seem to have no effect.

5. **Consider composability**: Design your wrappers so they could potentially be wrapped by even higher-level modules.

6. **Prefer reactive `defaults` over `update*Input()`**: To drive a base module parameter from your wrapper's state, pass a `reactive()` in `defaults`. Pushing values with `update*Input()` costs an extra render on every change. When you must update one of your *own* inputs from the server, call `freezeReactiveValue()` on it first so readers pause instead of rendering with the stale value. Freezing does not cover an input you rebuild with `renderUI()`, though. In those cases, resolve the value server-side and have the plot read that instead, as `setup_group_colors()` does for a colour picker (see the "Updating Your Own Inputs From the Server" section in `vignette("adding-a-new-module", package = "VizModules")` for more info).

See any of the existing modules for clear reference examples.
