---
title: "Quick Start with VizModules"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quick Start}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    eval = FALSE
)
```

## See the modules

Start with the hosted gallery to explore what each module can do:
<https://j-andrews7-VizModules.share.connect.posit.cloud/>

You can also run the same gallery locally from this package:

```{r run-gallery}
library(shiny)
shiny::runApp(system.file("apps/module-gallery", package = "VizModules"))
```

## Drop a module into your app

All modules follow the same pattern: `*InputsUI()` for controls, `*OutputUI()` for the plot, and `*Server()` for the logic. Here is a minimal `scatterPlot` example:

```{r scatter-example}
library(shiny)
library(VizModules)

ui <- fluidPage(
    sidebarLayout(
        sidebarPanel(
            dittoViz_ScatterPlotInputsUI("cars",
                mtcars,
                defaults = list(
                    x.by = "wt",
                    y.by = "mpg",
                    color.by = "cyl"
                )
            )
        ),
        mainPanel(dittoViz_ScatterPlotOutputUI("cars"))
    )
)

server <- function(input, output, session) {
    dittoViz_ScatterPlotServer("cars", data = reactive(mtcars))
}

shinyApp(ui, server)
```

Replace `scatterPlot` with any other module name (e.g., `AreaPlot`, `BarPlot`, `BoxPlot`, `linePlot`, `piePlot`, `volcanoPlot`) to swap the visualization.

## Set defaults and hide controls

- **Defaults**: Pass a named list to the `defaults` argument of `*InputsUI()` to pre-fill inputs. Names match the underlying plot function arguments (e.g., `defaults = list(fill.color = "steelblue")`).
- **Hide inputs**: Use `hide.inputs` in the server call to remove controls while still initializing their values. This is useful when your app sets certain parameters itself.
- **Hide tabs**: Use `hide.tabs` to remove whole groups of controls (e.g., `"Plotly"` or `"Legend/Scale"` in `scatterPlot`).

```{r hide-controls}
dittoViz_ScatterPlotServer(
    "cars",
    data = reactive(mtcars),
    hide.inputs = c("split.by", "rows.use"),
    hide.tabs = c("Plotly")
)
```

Hidden inputs and tabs still feed their values into the plot, so the module stays fully configured while exposing only what your users need.

## Which plot parameters are exposed?

Modules wrap plotting functions from **dittoViz** and **plotthis**. To see which arguments are available in a module:

1. Open the module input help page, e.g., `?dittoViz_ScatterPlotInputsUI` or `?AreaPlotInputsUI`. The **Details** section notes which arguments from the underlying plot function are wired through and any that are intentionally omitted.
2. Cross-reference the base plot documentation (`?dittoViz::scatterPlot`, `?plotthis::AreaPlot`, etc.). Input names in `defaults` line up with those function arguments when they are supported.

If an argument is listed as missing or non-functional in the module docs, it has been intentionally hidden because it does not round-trip well in the interactive Plotly output.
