---
title: "Quick Start with VizModules"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quick Start}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    eval = FALSE
)
```

## See the modules

Start with the hosted gallery to explore what each module can do:
<https://j-andrews7-VizModules.share.connect.posit.cloud/>

You can also run the same gallery locally from this package:

```{r run-gallery}
library(shiny)
shiny::runApp(system.file("apps/module-gallery", package = "VizModules"))
```

## Drop a module into your app

All modules follow the same pattern: `*InputsUI()` for controls, `*OutputUI()` for the plot, and `*Server()` for the logic.
Here is a minimal `scatterPlot` example:

```{r scatter-example}
library(VizModules)

ui <- fluidPage(
    sidebarLayout(
        sidebarPanel(
            dittoViz_scatterPlotInputsUI("cars",
                mtcars,
                defaults = list(
                    x.by = "wt",
                    y.by = "mpg",
                    color.by = "cyl"
                )
            )
        ),
        mainPanel(dittoViz_scatterPlotOutputUI("cars"))
    )
)

server <- function(input, output, session) {
    dittoViz_scatterPlotServer("cars", data = reactive(mtcars))
}

shinyApp(ui, server)
```


## Set defaults and hide controls

- **Defaults**: Pass a named list to the `defaults` argument of `*InputsUI()` to pre-fill inputs. Names match the underlying plot function arguments (e.g., `defaults = list(fill.color = "steelblue")`).
- **Hide inputs**: Use `hide.inputs` in the server call to remove controls while still initializing their values. This is useful when your app sets certain parameters itself.
- **Hide tabs**: Use `hide.tabs` to remove whole groups of controls (e.g., `"Plotly"` or `"Legend/Scale"` in `scatterPlot`).

```{r hide-controls}
dittoViz_scatterPlotServer(
    "cars",
    data = reactive(mtcars),
    hide.inputs = c("split.by", "rows.use"),
    hide.tabs = c("Plotly")
)
```

Hidden inputs and tabs still feed their values into the plot, so the module stays fully configured while exposing only what your users need.

## App factory with `createModuleApp()`

To enable simple, consistent testing of any module, we provide an app factory function that returns a full 
standalone app with data import, a filterable data table, and dataset switching for any module - 
`createModuleApp()`:

```{r create-module-app}
library(VizModules)

app <- createModuleApp(
    inputs_ui_fn = plotthis_BarPlotInputsUI,
    output_ui_fn = plotthis_BarPlotOutputUI,
    server_fn    = plotthis_BarPlotServer,
    data_list    = list("cars" = mtcars),
    title        = "My Bar Plot"
)
if (interactive()) runApp(app)
```

All built-in `*App()` convenience functions (e.g. `plotthis_BarPlotApp()`, `linePlotApp()`)
are thin wrappers around `createModuleApp()` with sensible default data. You can also pass
custom wrapper module functions to `createModuleApp()` for rapid prototyping.

## Export Summary Data: 

We provide `collect_source_data()` to assemble a compact record of the
plotted data, stats, UI inputs, and the rendered plot, and `create_source_download_handler()`
to turn that record into a downloadable `.zip`. `collect_source_data()`
requires a reactive plotly plot; the output summary can be optionally enriched by
both a stats reactive and a UI inputs reactive. `create_source_download_handler()` also accepts
a named list of summaries (one per plot), which is how the Panel Builder bundles
every plot on its canvas into a single download.

```{r data-summary}

if (interactive()) {
    library(shiny)
    library(plotly)

    ui <- fluidPage(
        plotlyOutput("plot"),
        downloadButton("download_summary", "Download Summary")
    )

    server <- function(input, output, session) {
        # A reactive plotly plot
        plot_reactive <- reactive({
            plot_ly(mtcars, x = ~wt, y = ~mpg, type = "scatter", mode = "markers")
        })

        # Optional: a reactive returning a stats data.frame
        stats_reactive <- reactive({
            data.frame(
                metric = c("mean_mpg", "sd_mpg"),
                value  = c(mean(mtcars$mpg), sd(mtcars$mpg))
            )
        })

        # Optional: capture all UI inputs as a named list
        AllInputs <- reactive({
            reactiveValuesToList(input)
        })

        output$plot <- renderPlotly(plot_reactive())

        # Assemble the summary, then wire up the download handler.
        plot_summary_reactive <- reactive({
            collect_source_data(
                plot_reactive   = plot_reactive,
                stats_reactive  = stats_reactive,
                inputs_reactive = AllInputs()
            )
        })

        output$download_summary <- create_source_download_handler(
            data_list     = plot_summary_reactive,
            filename_base = "my_plot_summary"
        )
    }

    shinyApp(ui, server)
}
```


## Which plot parameters are exposed?

Modules wrap plotting functions from **dittoViz**, **plotthis**, and native plotting functions. To see which arguments are available in a module:

1. Open the module input help page, e.g., `?dittoViz_scatterPlotInputsUI` or `?plotthis_AreaPlotInputsUI`. The **Details** section notes which arguments from the underlying plot function are wired through and any that are intentionally omitted.
2. Cross-reference the base plot documentation (`?dittoViz::scatterPlot`, `?plotthis::AreaPlot`, etc.). Input names in `defaults` line up with those function arguments when they are supported.

If an argument is listed as missing or non-functional in the module docs, it has been intentionally hidden because it does not round-trip well in the interactive Plotly output.
