---
title: "Using Custom Shiny Inputs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using Custom Shiny Inputs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    eval = FALSE
)
```

VizModules ships two custom Shiny inputs that go beyond standard `shiny::*Input` widgets: **multiColorPicker** and **multiDynamicInput**. This vignette covers `multiDynamicInput`, a general-purpose widget that lets users dynamically add and remove rows of heterogeneous inputs.

## Overview

`multiDynamicInput` renders a "+ Add" button and a container. Each time the user clicks Add, a new row appears with whatever fields you define. An "×" button on each row deletes it. The collected value is reported to `input[[inputId]]` as a named list of rows.

Unlike `insertUI`/`removeUI`, rows are managed entirely client-side via DOM template cloning — no server round-trips for add/delete.
 
## Basic Usage

```{r}
library(shiny)
library(VizModules)

ui <- fluidPage(
    multiDynamicInput(
        "layers",
        label = "Layers",
        row_spec = list(
            colour = list(type = "colour", args = list(value = "#000000")),
            size   = list(type = "numeric", args = list(value = 2, min = 0.5, max = 10)),
            style  = list(type = "select", args = list(choices = c("solid", "dashed", "dotted")))
        )
    ),
    verbatimTextOutput("value")
)

server <- function(input, output, session) {
    output$value <- renderPrint(input$layers)
}

shinyApp(ui, server)
```

## Key Arguments

| Argument | Description |
|----------|-------------|
| `inputId` | Shiny input id. The value is accessed via `input[[inputId]]`. |
| `label` | Label shown next to the + Add button. Also determines row naming (lowercased). |
| `row_spec` | Named list defining one row's fields. Each entry has `type` (or `fn`) and `args`. |
| `elements` | Optional list of pre-filled rows to show on startup. |
| `max_per_row` | Number of fields per visual line before wrapping (default 4). |
| `panel` | If `FALSE`, removes the border/background styling. |

## Defining Fields with `row_spec`

Each field in `row_spec` is a named list with:

- **`type`** — a shorthand alias: `"select"`, `"text"`, `"numeric"`, `"slider"`, `"checkbox"`, `"colour"` / `"color"`.
- **`fn`** — alternatively, any input constructor function (e.g. `shiny::dateInput`). Use this for inputs without a built-in alias.
- **`args`** — a list of arguments passed to the constructor (everything except `inputId`, which is auto-generated per row).

```{r}
# Using type aliases
row_spec = list(
    name = list(type = "text", args = list(placeholder = "Enter name")),
    age  = list(type = "numeric", args = list(value = 25, min = 0, max = 120))
)

# Using fn for a custom input
row_spec = list(
    date = list(fn = shiny::dateInput, args = list(label = "Date", value = Sys.Date()))
)
```

## Pre-filled Rows with `elements`

To show rows on startup with specific values:

```{r}
multiDynamicInput(
    "models",
    label = "Models",
    row_spec = list(
        model_type = list(type = "select", args = list(choices = c("lm", "glm"))),
        formula    = list(type = "text", args = list(placeholder = "y ~ x"))
    ),
    elements = list(
        models1 = list(model_type = "lm", formula = "revenue ~ units"),
        models2 = list(model_type = "glm", formula = "revenue ~ poly(units, 2)")
    )
)
```

The names (`models1`, `models2`) don't matter for input — they're auto-generated from the label at runtime. The values inside each element must match the `row_spec` field names.

## Reading the Value

The value in `input[[inputId]]` is a named list of rows:

```{r}
# With label = "Models" and two rows filled in:
input$models
#> list(
#>   models1 = list(model_type = "lm", formula = "revenue ~ units"),
#>   models2 = list(model_type = "glm", formula = "revenue ~ poly(units, 2)")
#> )
```

Row names are derived from the `label` argument (lowercased) plus an index: `models1`, `models2`, etc.

## Updating from the Server

Use `updateMultiDynamicInput()` to replace all rows or clear them:

```{r}
# Replace all rows
updateMultiDynamicInput(session, "models", elements = list(
    models1 = list(model_type = "lm", formula = "y ~ x")
))

# Clear all rows
updateMultiDynamicInput(session, "models", clear = TRUE)
```

## Backend-Specific Fields

When used with the model backend system (see `vignette("custom-model-lines")`), fields can be tagged with a `backend` attribute. These fields are hidden by default and only appear when the matching model type is selected:
 
```{r}
row_spec = list(
    model_type = list(type = "select", args = list(choices = c("lm", "drm"))),
    formula    = list(type = "text", args = list(placeholder = "y ~ x")),
    # Only visible when model_type == "drm"
    drc_fct    = list(type = "select", backend = "drm",
                      args = list(choices = c("LL.4", "LL.3")))
)
```

In practice you don't build this manually — `build_model_row_spec()` does it for you by collecting `fields` from all registered backends.

## Summary

- `multiDynamicInput()` — place in your UI
- `input[[inputId]]` — read the named list of rows in your server
- `updateMultiDynamicInput()` — set or clear rows from the server
- `build_model_row_spec()` — auto-build a row_spec from registered model backends
